The watershed transform in ITK - discussion and new developments

### 2006-06-09 article/Article.tex
Article and bibliography cleanup
Ran spell check over article, corrected grammar. Removed my big bibliography and replaced with a customized one.

darcs-hash:20060609003028-fafb9-24b197713b6513af0f4780eebc5d83f4af4004a2.gz

\begin{abstract}
\noindent
This report discusses various definitions and implementations of the
watershed transform as well as providing an introduction to some well
known techniques for applying the watershed transform to practical
problems. The discussion will be focussed on new and existing ITK
classes.
\end{abstract}

@@

Section \ref{sect:using-ws-itk} propose some applications of watershed
with ITK.

@@

\section{The watershed transform -- a brief history}
\label{sect:history}
The watershed transform was first proposed by Beucher and
Lantu\'{e}joul\cite{Beucher-Lantu-79} as a geophysical model of rain
falling on a terrain. The idea is that a raindrop falling on a surface
will trickle down the path of steepest descent to a minima. The set of
points on the surface that lead to the same minima is known as a
catchment basin and borders between catchment basins are watershed
lines\footnote{Major geographical features, like rivers or oceans, are
often considered as separated by watershed lines, usually along
mountain ranges.}. If an image is considered as a terrain and divided
into catchment basins then the hope is that each catchment basin would
contain an object of interest. The image representing the terrain will
be referred to as the {\em control surface} in this report.

Translating the model of rain falling on a terrain to an algorithm has
a number of problems\footnote{Although this is what the first ITK
implementation does.} which are discussed in Section
\ref{sect:rainfallmodel}. The alternative is to imagine a flooding
process, with each regional minima\footnote{A regional minima is a set
of pixels surrounded by pixels of higher value.} becoming the source
of a lake. The terrain is progressively flooded and lakes eventually
meet neighboring lakes. Virtual dams are constructed to keep the
neighboring lakes separate as the waterlevel rises. When the image
surface is completely fooded the virtual dams correspond to the
watershed lines\footnote{In 3D neighboring lakes meet at surfaces.}.

@@

Another major advance made soon after by Meyer and Beucher
\cite{Beucher93a,Meyer1994a}. Meyer defined the watershed transform in terms of {\em
topographic distances} through {\em lower complete} images\footnote{A lower complete
image is an image where there is no plateaus, excepted plateaus
which are regional minima. A common way to compute the lower completion of an
image, is to add the distance from the nearest lower pixel to the pixels of all
the plateaus. Lower completion have the property to produce the same image than
the input image if the input image is already lower complete. By lower completing
an image, we introduce a new order relation in the plateaus pixels which can be
used to solve the plateaus problem with watershed.}. These
definitions rigorously defined the behavior in the presence of
plateaus and lead to some new algorithms based on graph theory. Two
examples of these algorithms are {\em hill-climbing} and {\em image
integration} and will be discussed in more detail in this report.
Figures \ref{fig:cthead1-markers}, \ref{fig:circle} and \ref{fig:difficult-cases} illustrate
the possible problems with plateaus.

@@

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image - bright ring on constant background]{\includegraphics[scale=0.7]{circle_plateau}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1mf-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1mF-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1Mf-rgb}}
\subfigure[Morphological watershed]{\includegraphics[scale=0.7]{circle_rgb2}}
\subfigure[ITK watershed]{\includegraphics[scale=0.7]{circle_rgb1}}
\caption{Another example of plateau problem. The inside and outside of the ring have the same brightness, but the ring is arbitarily assigned to the outer region by the ITK watershed. It is split by the morphological watershed.\label{fig:circle}}
\end{center}
\end{figure}

@@

It should be evident that a number of arbitary choices have been made
here - for example it is impossible to break a flat region into
pieces, which would probably better fit the physical reality of
rainfall (consider a flat topped ridge which should perhaps be split
down the middle). What happens of there are multiple neighboring
regions with equal height - is the labelling decision now somewhat arbitary?

In general the flooding models avoid these dilemas and lead to more
elegant algorithms.

\section{Using the watershed tranform}
\label{sect:usingWT}
\subsection{Classical approach}
The simplest way of using the watershed is to preprocess the image we
want to segment so that the boundaries of our objects are bright (e.g
apply an edge detector) and compute the watershed transform of the
edge image. Watershed lines will correspond to the boundaries and our
problem will be solved. This is rarely useful in practice because
there are always more regional minima than there are objects, either
due to noise or natural variations in the object surfaces. Therefore,
while many watershed lines do lie on significant boundaries, there are
many that don't. Figure \ref{fig:overseg} illustrates the problem.

@@

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-rgb}}
\subfigure[Minima level = 10]{\includegraphics[scale=3.2]{level10-rgb}}
\subfigure[Minima level = 20]{\includegraphics[scale=3.2]{level20-rgb}}
\subfigure[Minima level = 30 and Minima level = 40]{\includegraphics[scale=3.2]{level30-rgb}}
\subfigure[Minima level = 50]{\includegraphics[scale=3.2]{level50-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markersMf-rgb}}
\caption{Different results of morphological watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image. All the watersheds were performed with $MarkWatershedLine = true$ and $FullyConnected = false$.\label{fig:artificial-itk}}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-itk-rgb}}
\subfigure[Level = 0.1]{\includegraphics[scale=3.2]{level10-itk-rgb}}
% \subfigure[Level = 0.2]{\includegraphics[scale=3.2]{level20-itk-rgb}}
\subfigure[Level = 0.4]{\includegraphics[scale=3.2]{level40-itk-rgb}}
\subfigure[Level = 0.63]{\includegraphics[scale=3.2]{level63-itk-rgb}}
\subfigure[Level = 0.65]{\includegraphics[scale=3.2]{level65-itk-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markers-itk-rgb}}
\caption{Different results of ITK's watershed on an articial image. The labeled image are overlayed to the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image.\label{fig:artificial-morpho}}
\end{center}
\end{figure}

@@

This can be achieved by performing a {\em reconstruction by erosion}
or {\em recursive conditional erosion} from the seleted regional
minima. This process is also called {\em minima imposition}. Figure
\ref{fig:recon} illustrates the result of applying a
reconstruction by erosion to a 1D image. The original image is shown
in blue and has regional minima A, B, C, D and E. The marker image
which is recursively eroded is shown in black, with a single zero at
regional minimum B. The red plateaus illustrate the parts of the
result image that are above the orginal. The resulting image has only
on regional minima that corresponds to B.
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.5]{recon}
\caption{Reconstruction by erosion from a single regional minimum.}
\label{fig:recon}
\end{center}
\end{figure}

@@

This procedure may not seem to offer much advantage -- after all how
are we going to decide which minima are useful to us? The key step to
using these ideas in practice is to realize that an arbitary mask
image can be used to create minima in an image in locations we want
simply by multiplying the image by the mask. Recursive erosion of the
original image using that mask then produces a topology to which we
can apply a watershed transform. The mask image is commonly called a
marker image.

@@

\begin{enumerate}
\item Generate a marker image. The marker image will be used to create 
one, possibly extended, minima inside each object of interest. A
marker image will always need to create at least two minima, and at
least one minima will probably be used to segment the background. The
marker image may be generated via interaction with an operator or by
various preprocessing steps.
\item \label{recipe:min}Impose the minima on the control surface. This 
might be done by multiplying by the marker image (or an inverted
version of it, depending on convention).
\item \label{recipe:erode}Recursively erode the control surface to eliminate all minima besides those introduced in the previous step. This produces a new control surface.
\item Apply the watershed tranform to the new surface.
\end{enumerate}

@@

\section{Algorithms}
\label{sect:algorithms}
The two that will be outlined here are known as {\em image
integration}\cite{Meyer1994a} and {\em hill climbing}
\cite{Beucher93a}, both of which are classed as ordered
algorithms and are examples of graph flooding approaches. These
algorithms implement the watershed tranform by topographic distance
(or approximate it) and are derived from Dijkstra-Moore minimal path
algorithms of graph theory.

\subsection{Ordered queues}
Both of these implementations depend on ordered queues to avoid the
need for explicit computation of lower complete images and recursive
erosion. An ordered queue is a heirarchical priority queue with two
levels. One level is defined by the pixel gray level, with darker
pixels having higher priority, while the second applies to pixels of
the same grey level and provides a FIFO ordering. The FIFO ordering
leads to a breadth first propogation across plateaus because the
ordering becomes related to distance from the regional minima. This is
the mechanism that avoids the to explicitly compute lower slope.

@@

Wavefronts are propogated from each minima, using the ordered queue
approach. As the wavefront reaches pixel the distance travelled by the
wavefront is compared to the distance stored in the distance image. If
the distance travelled by the new wavefront is less than the stored
value then that pixel is given the label associated with the
wavefront, otherwise the label is left unchanged.

@@

This algorithm take a grayscale image $I$ and a marker image $M$
which contains a set of markers with different labels. It can be
splitted in to steps:
\begin{enumerate}
  \item copy the marker image $M$ to the output image, and put
all the marker pixels $p$ with background neighbors in the
ordered queue, with the priority $I[p]$.
  \item get a pixel from the ordered queue at the highest
priority level, and propagate the label of $p$ to all the unlabeled
neighbors of $p$. The unlabeled neighbors $q$ are added in the
ordered queue with priority $I[q]$, if $I[q]$ is greater than
the priority level of $p$, or with the priority of $p$ otherwise.
\end{enumerate}

@@

\begin{figure}[htbp]
\centering
\small
\begin{pseudocode}[framebox]{BeucherWatershed}{I,M}
\COMMENT{initialization stage} \\
queue \GETS \emptyset \\
\FOREACH p \in M \DO
\BEGIN
  \COMMENT{copy the marker image to the output image}\\
  O[p] \GETS M[p] \\
  \COMMENT{put the edges of the marker in the queue}\\
  \IF M[p] \neq WSLABEL
  \THEN
  \BEGIN
    bgNeighbor \GETS \FALSE \\
    \FOREACH q \in \CALL{N}{p} \DO 
    \BEGIN
      \IF M[q] = WSLABEL
        \THEN bgNeighbor \GETS \TRUE \\
    \END\\
    \IF bgNeighbor \STMTNUM{5em}{algo:Beucher:bgNeighbor}
      \THEN
      queue[I[p]] \GETS queue[I[p]] \cup \{p\} \\
  \END
\END
\\
\\
\COMMENT{flooding stage} \\
\WHILE queue \neq \emptyset \DO
\BEGIN
  v \GETS \CALL{first}{queue}\\
  \WHILE queue[v] \neq \emptyset \DO
  \BEGIN
    p \GETS \CALL{first}{queue[v]} \\
    \FOREACH q \in \CALL{N}{p} \DO
    \BEGIN
    \IF O[q] = WSLABEL
    \THEN
    \BEGIN
      O[q] \GETS O[p] \\
      v \GETS \CALL{MAX}{I[q], v} \STMTNUM{5em}{algo:Beucher:max} \\
      queue[v] \GETS queue[v] \cup \{q\} \\
    \END
    \END
  \END
\END
\end{pseudocode}
\caption{Simple watershed algorithm.\label{BeucherAlgorithm} $FIRST(queue)$ extract the highest priority element from the queue. $N(p)$ return the list of p's neighbors. $MAX(a,b)$ returns the maximum of $a$, and $b$. $WSLABEL$ is the label used to define background and watershed pixels (usally $0$). The lower completion is done by the statement (\ref{algo:Beucher:max}). Determine if a marker pixel before adding it in the queue (\ref{algo:Beucher:bgNeighbor}) is not required, but helps to decrease memory usage in case of large markers.}
\end{figure}

@@

This algorithm take a grayscale image $I$ and a marker image $M$
which contains a set of markers with different labels. It can be
splitted in to steps:
\begin{enumerate}
  \item copy the marker image $M$ to the output image, and put
all the background pixels $p$ with markers in there neighbors in the
ordered queue, with the priority $I[p]$, and mark $p$ as already in
the queue. All the marker pixels are marked as already in the queue.
  \item get a pixel $p$ from the ordered queue at the highest
priority level. If the neighborhood contains only the same label
or watershed pixels, mark $p$ with that label. The neighbors $q$ 
of $p$ which are not already in the queue are added to the queue 
with priority $I[q]$, if $I[q]$ is greater than the priority level
of $p$, or with the priority of $p$ otherwise.
\end{enumerate}

@@

It should be noticed that:
\begin{itemize}
  \item this algorithm mark the watershed lines,
  \item the queue only contains the borders of the growing regions,
  \item this algorithm never propagates watershed
lines, and that the watershed lines it produces have a width of 1 pixel.
Also, the watershed lines form a connected obejct if we consider the
opposite connectivity than the one used in the algorithm: with a 2D
image the watershed lines are 8-connected if the algorithm is run
as 4-connected, and 4-connected if it is run as 8-connected.
\end{itemize}

\begin{figure}[htbp]
\centering
\small
\begin{pseudocode}[framebox]{MeyerWatershed}{I,M}
\COMMENT{initialization stage} \\
\FOREACH p \in M \DO
  S[p] \GETS \FALSE \\
queue \GETS \emptyset \\
\FOREACH p \in M \DO
\BEGIN
  \COMMENT{copy the marker image to the output image}\\
  O[p] \GETS M[p] \\
  \IF M[p] \neq WSLABEL
  \THEN
  \BEGIN
    S[p] \GETS \TRUE \\
    \COMMENT{add all the neighbors of $p$ with WSLABEL to the queue}\\
    \FOREACH q \in \CALL{N}{p} \DO
    \BEGIN
      \IF \neg S[q] \AND M[q] = WSLABEL 
      \THEN
      \BEGIN
        S[q] \GETS \TRUE \\
        queue[I[q]] \GETS queue[I[q]] \cup \{q\} \\
      \END
    \END
  \END
\END
\\
\\
\COMMENT{flooding stage} \\
\WHILE queue \neq \emptyset \DO
\BEGIN
  v \GETS \CALL{first}{queue}\\
  \WHILE queue[v] \neq \emptyset \DO
  \BEGIN
    p \GETS \CALL{first}{queue[v]} \\
    \COMMENT{found the label of $p$ and determine if it is a watershed pixel} \\
    label \GETS WSLABEL \\
    watershed \GETS \FALSE \\
    \FOREACH q \in \CALL{N}{p} \DO
    \BEGIN
      \IF O[q] \neq WSLABEL \AND \neg watershed
      \THEN
        \BEGIN
        \IF label \neq WSLABEL \AND O[q] \neq label \\
        \THEN
          watershed \GETS \TRUE \\
        \ELSE
          label \GETS O[q] \\
        \END
    \END \\
    \COMMENT{if $p$ is a watershed pixel, don't propagate this pixel} \\
    \IF \neg watershed
    \THEN
      \BEGIN
        \COMMENT{assign the label value to the current pixel} \\
        O[p] \GETS label \\
        \COMMENT{put all the neighbors of p in the queue} \\
        \FOREACH q \in \CALL{N}{p} \DO
        \BEGIN
          \IF \neg S[q]
          \THEN
            \BEGIN
            S[q] \GETS \TRUE \\
            h \GETS \CALL{MAX}{I[q], v} \STMTNUM{5em}{algo:Meyer:max} \\
            queue[h] \GETS queue[h] \cup \{q\} \\
            \END
        \END
      \END
  \END
\END
\end{pseudocode}
\caption{Watershed algorithm with watershed lines.\label{MeyerAlgorithm} $FIRST(queue)$ extract the highest priority element from the queue. $N(p)$ return the list of p's neighbors. $MAX(a,b)$ returns the maximum of $a$, and $b$. $WSLABEL$ is the label used to define background and watershed pixels (usally $0$). The lower completion is done by the statement (\ref{algo:Meyer:max})}
\end{figure}

@@

The watershed from markers can be implemented with the ITK watershed,
by using the technique of {\em minima imposition} exposed in Section
\ref{sect:ws-selected-minima}. Several steps are required:
\begin{itemize}
  \item Impose the minima to the input image, with the
\verb$MinimaImpositionImageFilter$
  \item Perform the watershed with the \verb$WatershedImageFilter$
  \item Restore the labels produced by the \verb$WatershedImageFilter$.
\verb$WatershedImageFilter$ produce arbitrary labels which doesn't match
the labels in the marker image. There is currently no filter able to
find the label correspondance in ITK, so this step must be performed by
hand with a code like the one before. The \verb$ChangeLabelImageFilter$
is well designed for this task:
\small \begin{verbatim}
filter->Update();

@@

This approach has several drawbacks:
\begin{itemize}
  \item it doesn't work if there some markers are not separated by background
pixels - they are considered as a single marker
  \item restoring the labels is not included in a filter and so break the
ITK pipeline model
  \item it's less efficient, as shown in Section \ref{sect:perf}
  \item it still gives poor results with plateaus. A lower completion
sould be performed before to also solve this problem\footnote{Lower
completion is not implemented yet in ITK}.
\end{itemize}

@@

We use a median image filter to remove some noise, and threshold the image
to get a binary image\footnote{The methods to remove the noise and binarize the
image are volontary kept simple for this example}.

>>>>

\begin{abstract}
\noindent
This report discusses various definitions and implementations of the
watershed transform as well as providing an introduction to some well
known techniques for applying the watershed transform to practical
problems. The discussion will be focused on new and existing ITK
classes.
\end{abstract}

@@

Section \ref{sect:using-ws-itk} proposes some applications of watershed
with ITK.

@@

\section{The watershed transform -- a brief history}
\label{sect:history}
The watershed transform was first proposed by Beucher and
Lantu\'{e}joul\cite{Beucher-Lantu-79} as a geophysical model of rain
falling on a terrain. The idea is that a raindrop falling on a surface
will trickle down the path of steepest descent to a minimum. The set of
points on the surface that lead to the same minimum is known as a
catchment basin and borders between catchment basins are watershed
lines\footnote{Major geographical features, like rivers or oceans, are
often considered as separated by watershed lines, usually along
mountain ranges.}. If an image is considered as a terrain and divided
into catchment basins then the hope is that each catchment basin would
contain an object of interest. The image representing the terrain will
be referred to as the {\em control surface} in this report.

Translating the model of rain falling on a terrain to an algorithm has
a number of problems\footnote{Although this is what the first ITK
implementation does.} which are discussed in Section
\ref{sect:rainfallmodel}. The alternative is to imagine a flooding
process, with each regional minima\footnote{A regional minima is a set
of pixels surrounded by pixels of higher value.} becoming the source
of a lake. The terrain is progressively flooded and lakes eventually
meet neighboring lakes. Virtual dams are constructed to keep the
neighboring lakes separate as the water level rises. When the image
surface is completely flooded the virtual dams correspond to the
watershed lines\footnote{In 3D neighboring lakes meet at surfaces.}.

@@

Another major advance made soon after by Meyer and Beucher
\cite{Beucher93a,Meyer1994a}. Meyer defined the watershed transform in terms of {\em
topographic distances} through {\em lower complete} images\footnote{A
lower complete image is an image where the only plateaus are regional
minima. A common way to compute the lower completion of an image, is
to add the distance from the nearest lower pixel to the pixels of all
the plateaus. Lower completion have the property to produce the same
image than the input image if the input image is already lower
complete. By lower completing an image, we introduce a new order
relation in the plateaus pixels which can be used to solve the
plateaus problem with watershed.}. These definitions rigorously
defined the behavior in the presence of plateaus and lead to some new
algorithms based on graph theory. Two examples of these algorithms are
{\em hill-climbing} and {\em image integration} and will be discussed
in more detail in this report.  Figures \ref{fig:cthead1-markers},
\ref{fig:circle} and \ref{fig:difficult-cases} illustrate the possible
problems with plateaus.

@@

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image - bright ring on constant background]{\includegraphics[scale=0.7]{circle_plateau}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1mf-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1mF-rgb}}
% \subfigure[]{\includegraphics[scale=0.7]{cthead1Mf-rgb}}
\subfigure[Morphological watershed]{\includegraphics[scale=0.7]{circle_rgb2}}
\subfigure[ITK watershed]{\includegraphics[scale=0.7]{circle_rgb1}}
\caption{Another example of plateau problem. The inside and outside of the ring have the same brightness, but the ring is arbitrarily assigned to the outer region by the ITK watershed. It is split by the morphological watershed.\label{fig:circle}}
\end{center}
\end{figure}

@@

It should be evident that a number of arbitrary choices have been made
here - for example it is impossible to break a flat region into
pieces, which would probably better fit the physical reality of
rainfall (consider a flat topped ridge which should perhaps be split
down the middle). What happens of there are multiple neighboring
regions with equal height - is the labelling decision now somewhat arbitrary?

In general the flooding models avoid these dilemmas and lead to more
elegant algorithms.

\section{Using the watershed transform}
\label{sect:usingWT}
\subsection{Classical approach}
The simplest way of using the watershed is to preprocess the image we
want to segment so that the boundaries of our objects are bright (e.g
apply an edge detector) and compute the watershed transform of the
edge image. Watershed lines will correspond to the boundaries and our
problem will be solved. This is rarely useful in practice because
there are always more regional minima than there are objects, either
due to noise or natural variations in the object surfaces. Therefore,
while many watershed lines do lie on significant boundaries, there are
many that don't. Figure \ref{fig:overseg} illustrates the problem.

@@

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-rgb}}
\subfigure[Minima level = 10]{\includegraphics[scale=3.2]{level10-rgb}}
\subfigure[Minima level = 20]{\includegraphics[scale=3.2]{level20-rgb}}
\subfigure[Minima level = 30 and Minima level = 40]{\includegraphics[scale=3.2]{level30-rgb}}
\subfigure[Minima level = 50]{\includegraphics[scale=3.2]{level50-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markersMf-rgb}}
\caption{Different results of morphological watershed on an artificial image. The labeled image are overlayed on the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image. All the watersheds were performed with $MarkWatershedLine = true$ and $FullyConnected = false$.\label{fig:artificial-itk}}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\subfigure[Input image]{\includegraphics[scale=3.2]{level}}
\subfigure[No minima level]{\includegraphics[scale=3.2]{level00-itk-rgb}}
\subfigure[Level = 0.1]{\includegraphics[scale=3.2]{level10-itk-rgb}}
% \subfigure[Level = 0.2]{\includegraphics[scale=3.2]{level20-itk-rgb}}
\subfigure[Level = 0.4]{\includegraphics[scale=3.2]{level40-itk-rgb}}
\subfigure[Level = 0.63]{\includegraphics[scale=3.2]{level63-itk-rgb}}
\subfigure[Level = 0.65]{\includegraphics[scale=3.2]{level65-itk-rgb}}
\subfigure[markers]{\includegraphics[scale=3.2]{level-markers-rgb}}
\subfigure[Watershed from markers]{\includegraphics[scale=3.2]{level-markers-itk-rgb}}
\caption{Different results of ITK's watershed on an artificial image. The labeled image are overlayed on the input image with {\em LabelOverlayImageFilter}. From $(b)$ to $(f)$, an example of reduction of the classical watershed over segmentation by minima dynamic filtering. $(g)$ and $(h)$ show a marker defined by a user, and the result over the input image.\label{fig:artificial-morpho}}
\end{center}
\end{figure}

@@

This can be achieved by performing a {\em reconstruction by erosion}
or {\em recursive conditional erosion} from the selected regional
minima. This process is also called {\em minima imposition}. Figure
\ref{fig:recon} illustrates the result of applying a
reconstruction by erosion to a 1D image. The original image is shown
in blue and has regional minima A, B, C, D and E. The marker image
which is recursively eroded is shown in black, with a single zero at
regional minimum B. The red plateaus illustrate the parts of the
result image that are above the original. The resulting image has only
on regional minima that corresponds to B.
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.5]{recon}
\caption{Reconstruction by erosion from a single regional minimum.}
\label{fig:recon}
\end{center}
\end{figure}

@@

This procedure may not seem to offer much advantage -- after all how
are we going to decide which minima are useful to us? The key step to
using these ideas in practice is to realize that an arbitrary mask
image can be used to create minima in an image in locations we want
simply by multiplying the image by the mask. Recursive erosion of the
original image using that mask then produces a topology to which we
can apply a watershed transform. The mask image is commonly called a
marker image.

@@

\begin{enumerate}
\item Generate a marker image. The marker image will be used to create 
one, possibly extended, minima inside each object of interest. A
marker image will always need to create at least two minima, and at
least one minima will probably be used to segment the background. The
marker image may be generated via interaction with an operator or by
various preprocessing steps.
\item \label{recipe:min}Impose the minima on the control surface. This 
might be done by multiplying by the marker image (or an inverted
version of it, depending on convention).
\item \label{recipe:erode}Recursively erode the control surface to eliminate all minima besides those introduced in the previous step. This produces a new control surface.
\item Apply the watershed transform to the new surface.
\end{enumerate}

@@

\section{Algorithms}
\label{sect:algorithms}
The two that will be outlined here are known as {\em image
integration}\cite{Meyer1994a} and {\em hill climbing}
\cite{Beucher93a}, both of which are classed as ordered
algorithms and are examples of graph flooding approaches. These
algorithms implement the watershed transform by topographic distance
(or approximate it) and are derived from Dijkstra-Moore minimal path
algorithms of graph theory.

\subsection{Ordered queues}
Both of these implementations depend on ordered queues to avoid the
need for explicit computation of lower complete images and recursive
erosion. An ordered queue is a hierarchical priority queue with two
levels. One level is defined by the pixel gray level, with darker
pixels having higher priority, while the second applies to pixels of
the same grey level and provides a FIFO ordering. The FIFO ordering
leads to a breadth first propagation across plateaus because the
ordering becomes related to distance from the regional minima. This is
the mechanism that avoids the to explicitly compute lower slope.

@@

Wavefronts are propagated from each minima, using the ordered queue
approach. As the wavefront reaches pixel the distance travelled by the
wavefront is compared to the distance stored in the distance image. If
the distance travelled by the new wavefront is less than the stored
value then that pixel is given the label associated with the
wavefront, otherwise the label is left unchanged.

@@

This algorithm take a gray scale image $I$ and a marker image $M$
which contains a set of markers with different labels. It can be
split into two steps:
\begin{enumerate}
  \item copy the marker image $M$ to the output image, and put
all the marker pixels $p$ with background neighbors in the
ordered queue, with the priority $I[p]$.
  \item get a pixel from the ordered queue at the highest
priority level, and propagate the label of $p$ to all the unlabeled
neighbors of $p$. The unlabeled neighbors $q$ are added in the
ordered queue with priority $I[q]$, if $I[q]$ is greater than
the priority level of $p$, or with the priority of $p$ otherwise.
\end{enumerate}

@@

\begin{figure}[htbp]
\centering
\small
\begin{pseudocode}[framebox]{BeucherWatershed}{I,M}
\COMMENT{initialization stage} \\
queue \GETS \emptyset \\
\FOREACH p \in M \DO
\BEGIN
  \COMMENT{copy the marker image to the output image}\\
  O[p] \GETS M[p] \\
  \COMMENT{put the edges of the marker in the queue}\\
  \IF M[p] \neq WSLABEL
  \THEN
  \BEGIN
    bgNeighbor \GETS \FALSE \\
    \FOREACH q \in \CALL{N}{p} \DO 
    \BEGIN
      \IF M[q] = WSLABEL
        \THEN bgNeighbor \GETS \TRUE \\
    \END\\
    \IF bgNeighbor \STMTNUM{5em}{algo:Beucher:bgNeighbor}
      \THEN
      queue[I[p]] \GETS queue[I[p]] \cup \{p\} \\
  \END
\END
\\
\\
\COMMENT{flooding stage} \\
\WHILE queue \neq \emptyset \DO
\BEGIN
  v \GETS \CALL{first}{queue}\\
  \WHILE queue[v] \neq \emptyset \DO
  \BEGIN
    p \GETS \CALL{first}{queue[v]} \\
    \FOREACH q \in \CALL{N}{p} \DO
    \BEGIN
    \IF O[q] = WSLABEL
    \THEN
    \BEGIN
      O[q] \GETS O[p] \\
      v \GETS \CALL{MAX}{I[q], v} \STMTNUM{5em}{algo:Beucher:max} \\
      queue[v] \GETS queue[v] \cup \{q\} \\
    \END
    \END
  \END
\END
\end{pseudocode}
\caption{Simple watershed algorithm.\label{BeucherAlgorithm} $FIRST(queue)$ extract the highest priority element from the queue. $N(p)$ return the list of p's neighbors. $MAX(a,b)$ returns the maximum of $a$, and $b$. $WSLABEL$ is the label used to define background and watershed pixels (usually $0$). The lower completion is done by the statement (\ref{algo:Beucher:max}). Determine if a marker pixel before adding it in the queue (\ref{algo:Beucher:bgNeighbor}) is not required, but helps to decrease memory usage in case of large markers.}
\end{figure}

@@

This algorithm take a gray scale image $I$ and a marker image $M$
which contains a set of markers with different labels. It can be
split into two steps:
\begin{enumerate}
  \item copy the marker image $M$ to the output image, and put
all the background pixels $p$ with markers in there neighbors in the
ordered queue, with the priority $I[p]$, and mark $p$ as already in
the queue. All the marker pixels are marked as already in the queue.
  \item get a pixel $p$ from the ordered queue at the highest
priority level. If the neighborhood contains only the same label
or watershed pixels, mark $p$ with that label. The neighbors $q$ 
of $p$ which are not already in the queue are added to the queue 
with priority $I[q]$, if $I[q]$ is greater than the priority level
of $p$, or with the priority of $p$ otherwise.
\end{enumerate}

@@

It should be noticed that:
\begin{itemize}
  \item this algorithm mark the watershed lines,
  \item the queue only contains the borders of the growing regions,
  \item this algorithm never propagates watershed
lines, and that the watershed lines it produces have a width of 1 pixel.
Also, the watershed lines form a connected object if we consider the
opposite connectivity than the one used in the algorithm: with a 2D
image the watershed lines are 8-connected if the algorithm is run
as 4-connected, and 4-connected if it is run as 8-connected.
\end{itemize}

\begin{figure}[htbp]
\centering
\small
\begin{pseudocode}[framebox]{MeyerWatershed}{I,M}
\COMMENT{initialization stage} \\
\FOREACH p \in M \DO
  S[p] \GETS \FALSE \\
queue \GETS \emptyset \\
\FOREACH p \in M \DO
\BEGIN
  \COMMENT{copy the marker image to the output image}\\
  O[p] \GETS M[p] \\
  \IF M[p] \neq WSLABEL
  \THEN
  \BEGIN
    S[p] \GETS \TRUE \\
    \COMMENT{add all the neighbors of $p$ with WSLABEL to the queue}\\
    \FOREACH q \in \CALL{N}{p} \DO
    \BEGIN
      \IF \neg S[q] \AND M[q] = WSLABEL 
      \THEN
      \BEGIN
        S[q] \GETS \TRUE \\
        queue[I[q]] \GETS queue[I[q]] \cup \{q\} \\
      \END
    \END
  \END
\END
\\
\\
\COMMENT{flooding stage} \\
\WHILE queue \neq \emptyset \DO
\BEGIN
  v \GETS \CALL{first}{queue}\\
  \WHILE queue[v] \neq \emptyset \DO
  \BEGIN
    p \GETS \CALL{first}{queue[v]} \\
    \COMMENT{found the label of $p$ and determine if it is a watershed pixel} \\
    label \GETS WSLABEL \\
    watershed \GETS \FALSE \\
    \FOREACH q \in \CALL{N}{p} \DO
    \BEGIN
      \IF O[q] \neq WSLABEL \AND \neg watershed
      \THEN
        \BEGIN
        \IF label \neq WSLABEL \AND O[q] \neq label \\
        \THEN
          watershed \GETS \TRUE \\
        \ELSE
          label \GETS O[q] \\
        \END
    \END \\
    \COMMENT{if $p$ is a watershed pixel, don't propagate this pixel} \\
    \IF \neg watershed
    \THEN
      \BEGIN
        \COMMENT{assign the label value to the current pixel} \\
        O[p] \GETS label \\
        \COMMENT{put all the neighbors of p in the queue} \\
        \FOREACH q \in \CALL{N}{p} \DO
        \BEGIN
          \IF \neg S[q]
          \THEN
            \BEGIN
            S[q] \GETS \TRUE \\
            h \GETS \CALL{MAX}{I[q], v} \STMTNUM{5em}{algo:Meyer:max} \\
            queue[h] \GETS queue[h] \cup \{q\} \\
            \END
        \END
      \END
  \END
\END
\end{pseudocode}
\caption{Watershed algorithm with watershed lines.\label{MeyerAlgorithm} $FIRST(queue)$ extract the highest priority element from the queue. $N(p)$ return the list of p's neighbors. $MAX(a,b)$ returns the maximum of $a$, and $b$. $WSLABEL$ is the label used to define background and watershed pixels (usually $0$). The lower completion is done by the statement (\ref{algo:Meyer:max})}
\end{figure}

@@

The watershed from markers can be implemented with the ITK watershed,
by using the technique of {\em minima imposition} exposed in Section
\ref{sect:ws-selected-minima}. Several steps are required:
\begin{itemize}
  \item Impose the minima to the input image, with the
\verb$MinimaImpositionImageFilter$
  \item Perform the watershed with the \verb$WatershedImageFilter$
  \item Restore the labels produced by the \verb$WatershedImageFilter$.
\verb$WatershedImageFilter$ produce arbitrary labels which doesn't match
the labels in the marker image. There is currently no filter able to
find the label correspondence in ITK, so this step must be performed by
hand with a code like the one before. The \verb$ChangeLabelImageFilter$
is well designed for this task:
\small \begin{verbatim}
filter->Update();

@@

This approach has several drawbacks:
\begin{itemize}
  \item it doesn't work if there some markers are not separated by background
pixels - they are considered as a single marker
  \item restoring the labels is not included in a filter and so break the
ITK pipeline model
  \item it's less efficient, as shown in Section \ref{sect:perf}
  \item it still gives poor results with plateaus. A lower completion
souled be performed before to also solve this problem\footnote{Lower
completion is not implemented yet in ITK}.
\end{itemize}

@@

We use a median image filter to remove some noise, and threshold the image
to get a binary image\footnote{The methods to remove the noise and binarize the
image are kept simple for this example}.
